Clique and Constraint Models for Maximum Common (Connected) Subgraph Problems

### 2016-06-08 paper/paper.tex
Why 33 for labels
Betterer wording

Then, in \cref{mccs}, we consider the MCCS problem. For the CP approach, we may add a global
connectedness constraint to the model. Alternatively, we may use a special branching rule
\cite{DBLP:conf/mco/VismaraV08} to grow connected subgraphs only. These two techniques may
be combined, and we experimentally show that the best results are obtained when combining them.
When solving the MCCS problem with a clique-based approach, neither technique seems directly viable
with an association graph encoding.  However, we show that it is possible to adapt the combined
branching and bounding rule used by modern clique algorithms to maintain connectedness during
search.  We compare the clique-based approach with the best CP variant for MCCS, and we show
that it outperforms CP on labelled graphs, whereas it is outperformed by CP on unlabelled graphs.

@@

McGregor \cite{McGreg82} proposed a branch and bound approach: each branch of the search tree
corresponds to the matching of two vertices, and a bounding function evaluates the number of
vertices that still may be matched so that the current branch is pruned as soon as this bound
becomes lower than the size of the largest known common subgraph. CP approaches may be viewed as
enhancements of this branch and bound algorithm.

@@

This CP model is improved by Ndiaye and Solnon \cite{DBLP:conf/cp/NdiayeS11} by replacing binary
difference constraints with a soft global allDifferent constraint which maximizes the number of
$x_u$ variables that are assigned to values different from $\bot$, while ensuring they are all
different when they are not assigned to $\bot$. They find that the combination ``MAC+Bound'' (resp.
``FC+Bound'') obtains the best results on labelled (resp. unlabelled) graphs and outperforms the
two previous approaches. The combination ``MAC+Bound'' maintains arc consistency~\cite{sabi94} of
edge constraints, whereas the combination ``FC+Bound'' simply performs forward-checking on these
constraints. In both combinations, the ``Bound'' filtering checks whether it is possible to assign
distinct values to enough $x_u$ variables to surpass the best cost found so far (it is a weaker
version of GAC(\emph{softAllDiff}) \cite{peti01} which computes the maximum number of variables that
can be assigned distinct values).

@@

We consider a randomly generated database
\cite{DBLP:journals/prl/SantoFSV03,DBLP:journals/jgaa/ConteFV07} commonly used for benchmarking
maximum common subgraph problems.  The dataset contains different classes of graphs: randomly
connected graphs with different densities; 2D, 3D, and 4D regular and irregular meshes; regular
bounded valence graphs, and irregular bounded valence graphs.  For each pair of graphs, there are 3
different labellings such that the number of different labels is equal to 33\%, 50\% or 75\% of the
number of vertices. In this paper, we report experiments with unlabelled graphs (labels are
ignored), and with 33\% labellings. For unlabelled graphs, we select the 27,500 graph pairs where the
number of vertices in each graph is no more than 35; for labelled graphs, which we find less
computationally challenging, we select all 81,400 graph pairs, to include graphs with up to 100
vertices.

@@

A more traditional CP approach would be to express connectedness as a conventional constraint. For
example, CP(Graph) \cite{DBLP:conf/cp/DoomsDD05} introduces graph domain variables and enforces
connectivity via the \emph{reachable} constraint, ensuring that there is a path from a specified
vertex to a specified set of vertices. One such constraint is then posted for each of the vertices
in the graph and encodes the transitive closure of the graph. Brown et al.\ \cite{Brown:2005}
explored the use of constraint programming in the generation of connected graphs with specified
degree sequences.  Two constraints were combined: the graphical constraint (a backtrackable
implementation of the Havel-Hakimi algorithm), and a connectivity constraint implemented using sets
of vertices, where vertex sets $A$ and $B$ are combined when there exists a pair of vertices $v \in
A$ and $w \in B$ and an edge $(v,w) \in E$. Residual degree counts are maintained on components and
vertices to enforce graphicality and connectivity. Prosser and Unsworth
\cite{DBLP:conf/ecai/ProsserU06} proposed a connectivity constraint for connected graph generation
where decision variables are edges (the search process accepts and rejects edges). The constraint
employed depth first search to maintain the set of tree edges and back edges, associating path
counters on these edges. The counters were then used to detect the existence of cut-edges and
protects these by forcing edges; the complexity of this constraint is cubic in the number of
vertices.

@@

During the search, each time a variable is assigned to $\bot$, we remove the corresponding vertex
from $S$ and perform a new traversal of $G[S]$ starting from the initial vertex $u$. For each vertex
$w\in S$ that is not visited by the traversal, we remove $w$ from $S$ and assign $x_w$ to $\bot$.
Also, each time a value is removed from a domain so that this value no longer belongs to any domain,
we remove it from $T$, and perform a new traversal of $H[T]$ starting from the initial vertex $v$.
For each vertex $w$ that is not visited by the traversal, we remove $w$ from $T$, and remove $w$
from all domains to which it belongs.

@@

Our $\FuncSty{colour}$ procedure is a simple greedy sequential colouring. We use the bit-parallel
algorithm introduced by San Segundo \textit{et al.}\ \cite{DBLP:journals/cor/SegundoRJ11}, with the
$k_{\mathit{min}}$ parameter set to $0$, so we do not present it here. We use a simple static degree
ordering; other initial vertex orderings have been considered on general clique problems
\cite{DBLP:journals/algorithms/Prosser12,DBLP:conf/lion/SegundoLB14}, and it is possible that
special properties of the association graph could be exploited in this step (for example, it is
always possible to colour the initial association graph using
$\operatorname{min}(\left|\operatorname{V}(G_1)\right|, \left|\operatorname{V}(G_2)\right|)$
colours, but with certain vertex orderings, a greedy sequential colouring will sometimes use
many more colours).

>>>>

Then, in \cref{mccs}, we consider the MCCS problem. For the CP approach, we may add a global
connectedness constraint to the model. Alternatively, we may use a special branching rule
\cite{DBLP:conf/mco/VismaraV08} to grow connected subgraphs only. These two techniques may be
combined, and we experimentally show that the best results are in fact obtained when combining them.
When solving the MCCS problem with a clique-based approach, neither technique seems directly viable
with an association graph encoding.  However, we show that it is possible to adapt the combined
branching and bounding rule used by modern clique algorithms to maintain connectedness during
search.  We compare the clique-based approach with the best CP variant for MCCS, and we show that it
outperforms CP on labelled graphs, whereas it is outperformed by CP on unlabelled graphs.

@@

McGregor \cite{McGreg82} proposed a branch and bound algorithm: each branch of the search tree
corresponds to the matching of two vertices, and a bounding function evaluates the number of
vertices that still may be matched so that the current branch is pruned as soon as this bound
becomes lower than the size of the largest known common subgraph. CP approaches may be viewed as
enhancements of this branch and bound algorithm.

@@

This CP model was improved by Ndiaye and Solnon \cite{DBLP:conf/cp/NdiayeS11} by replacing binary
difference constraints with a soft global allDifferent constraint which maximizes the number of
$x_u$ variables that are assigned to values different from $\bot$, while ensuring they are all
different when they are not assigned to $\bot$. They find that the combination ``MAC+Bound'' (resp.
``FC+Bound'') obtains the best results on labelled (resp. unlabelled) graphs and outperforms the
two previous approaches. The combination ``MAC+Bound'' maintains arc consistency~\cite{sabi94} of
edge constraints, whereas the combination ``FC+Bound'' simply performs forward-checking on these
constraints. In both combinations, the ``Bound'' filtering checks whether it is possible to assign
distinct values to enough $x_u$ variables to surpass the best cost found so far---it is a weaker
version of GAC(\emph{softAllDiff}) \cite{peti01} which computes the maximum number of variables that
can be assigned distinct values.

@@

We consider a randomly generated database
\cite{DBLP:journals/prl/SantoFSV03,DBLP:journals/jgaa/ConteFV07} commonly used for benchmarking
maximum common subgraph problems.  The dataset contains different classes of graphs: randomly
connected graphs with different densities; 2D, 3D, and 4D regular and irregular meshes; regular
bounded valence graphs, and irregular bounded valence graphs.  For each pair of graphs, there are 3
different labellings such that the number of different labels is equal to 33\%, 50\% or 75\% of the
number of vertices. In this paper, we report experiments with unlabelled graphs (labels are
ignored), and with 33\% labellings (the problem becomes very easy with larger numbers of labels).
For unlabelled graphs, we select the 27,500 graph pairs where the number of vertices in each graph
is no more than 35; for labelled graphs, which we find less computationally challenging, we select
all 81,400 graph pairs, to include graphs with up to 100 vertices.

@@

A more traditional CP approach would be to express connectedness as a conventional constraint. For
example, CP(Graph) \cite{DBLP:conf/cp/DoomsDD05} introduces graph domain variables and enforces
connectivity via the \emph{reachable} constraint, ensuring that there is a path from a specified
vertex to a specified set of vertices. One such constraint could be posted for each of the vertices
in the graph, encoding the transitive closure of the graph. Brown et al.\ \cite{Brown:2005}
explored the use of constraint programming in the generation of connected graphs with specified
degree sequences.  Two constraints were combined: the graphical constraint (a backtrackable
implementation of the Havel-Hakimi algorithm), and a connectivity constraint implemented using sets
of vertices, where vertex sets $A$ and $B$ are combined when there exists a pair of vertices $v \in
A$ and $w \in B$ and an edge $(v,w) \in E$. Residual degree counts are maintained on components and
vertices to enforce graphicality and connectivity. Prosser and Unsworth
\cite{DBLP:conf/ecai/ProsserU06} proposed a connectivity constraint for connected graph generation
where decision variables are edges (the search process accepts and rejects edges). The constraint
employed depth first search to maintain the set of tree edges and back edges, associating path
counters on these edges. The counters were then used to detect the existence of cut-edges and
protects these by forcing edges; the complexity of this constraint is cubic in the number of
vertices.

@@

During search, each time a variable is assigned to $\bot$, we remove the corresponding vertex
from $S$ and perform a new traversal of $G[S]$ starting from the initial vertex $u$. For each vertex
$w\in S$ that is not visited by the traversal, we remove $w$ from $S$ and assign $x_w$ to $\bot$.
Also, each time a value is removed from a domain so that this value no longer belongs to any domain,
we remove it from $T$, and perform a new traversal of $H[T]$ starting from the initial vertex $v$.
For each vertex $w$ that is not visited by the traversal, we remove $w$ from $T$, and remove $w$
from all domains to which it belongs.

@@

Our $\FuncSty{colour}$ procedure is a simple greedy sequential colouring. We use the bit-parallel
algorithm introduced by San Segundo \textit{et al.}\ \cite{DBLP:journals/cor/SegundoRJ11}, with the
$k_{\mathit{min}}$ parameter set to $0$, so we do not describe it here. We use a simple static degree
ordering; other initial vertex orderings have been considered on general clique problems
\cite{DBLP:journals/algorithms/Prosser12,DBLP:conf/lion/SegundoLB14}, and it is possible that
special properties of the association graph could be exploited in this step (for example, it is
always possible to colour the initial association graph using
$\operatorname{min}(\left|\operatorname{V}(G_1)\right|, \left|\operatorname{V}(G_2)\right|)$
colours, but with certain vertex orderings, a greedy sequential colouring will sometimes use
many more colours).
